---
title: "Coming from SAS"
vignette: >
  %\VignetteIndexEntry{Coming from SAS}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hvtiRdatasets)
```

## Who this is for

You know the `tp.bd.*` and `tp.vars.*` templates. You now have to read and
write R. This guide will not teach you R. Its job is to make what you already
know transferable, and to warn you about the handful of places where a
reasonable translation from SAS silently produces a *different, plausible*
number rather than an error.

Read the traps section even if you skip everything else.

## The model shift

A SAS program mutates a dataset through a sequence of steps, and the program
itself is the record of what happened. If an analyst commented out a block with
`%macro skip`, that decision lives only in their copy of the script.

R passes values through functions, and the study's `study.yaml` is the record.
Enabling a step is a change to a committed configuration file, not an
uncommented block. This is the whole reason the port is worth doing: what used
to be invisible becomes reviewable.

## Idiom translation

```{r}
#| eval: false
# Functions marked (S1)-(S3) arrive in later slices.
```

| SAS | R |
|---|---|
| `%include "<dbcreds>.sas"` + `CONNECT TO ODBC` | `dw_connect()` (S1) |
| `PROC SQL; SELECT ... FROM connection to ODBC` | `dw_pull(config, conn)` (S1) |
| `libname library "&STUDY/datasets"` | paths in `study.yaml` |
| `%macro skip; ... %mend skip;` | `modules:` / `derive:` toggles in `study.yaml` |
| `%vars(in=built, out=built, transf=1)` | `derive_vars(built, config)` (S3) |
| `data x; set y; ... run;` | a `dplyr` pipeline |
| `PROC CONTENTS` | `hvtiRutilities::data_dictionary()` |
| `label x = 'Age at surgery';` | `labelled` + `hvtiRutilities::label_map()` |
| `first.id` / `last.id` | `group_by()` + `slice_head()` / `slice_tail()` |
| `libname out xport` then `read.xport()` | parquet, or just return the object |

## The traps

Each of these produces a wrong number rather than an error. They are the reason
`compare_built()` exists.

### Missing sorts low

In SAS a missing numeric orders *below every number*. So this marks a patient
with unknown BMI as underweight:

```
if bmi < 18.5 then underweight = 1;
```

R does not:

```{r}
bmi <- c(31.2, NA, 17.0)
bmi < 18.5
```

`NA` propagates instead of comparing as small. R is right and SAS is wrong, so
a faithful port will *fail* equivalence here. That is what the
`intentional_divergence` resolution in `equivalence_signoff.yaml` is for. Do
not reproduce the SAS behaviour to make the comparison green.

### Character comparison is blank-padded in SAS

SAS pads character values to their declared length, so `'Smith'` and
`'Smith   '` compare equal. In R they do not:

```{r}
"Smith" == "Smith   "
trimws("Smith   ") == "Smith"
```

This bites hardest on merges keyed by a character id. `compare_built()` trims
before comparing, for exactly this reason.

### SAS has no missing character value

R distinguishes `NA` from `""`. SAS does not — a missing character **is** the
empty string. Read a SAS dataset and every missing character comes back as
`""`:

```{r}
f <- system.file("extdata", "oracle_small.sas7bdat",
                 package = "hvtiRdatasets")
surgeon <- haven::read_sas(f)$surgeon
surgeon                 # the fourth value was written as NA
is.na(surgeon)          # ... and is not NA any more
```

So a faithful R rebuild holding `NA` disagrees with the oracle holding `""` on
every missing value. `compare_built()` folds `""` to `NA` for character columns
so this does not flood the report with false differences.

The consequence for your own code is sharper: **never test a SAS-sourced
character for `NA`**. `is.na(surgeon)` is `FALSE` even where the value is
missing. Test for `""`, or convert on read.

### `MERGE` is not a join

A SAS data-step `MERGE` on a non-unique BY key does not error. It produces
undefined results. This is not hypothetical — from the revision history of
`tp.bd.data.master.sas`:

> 07/03/23: Changed the join logic for the FUP dataset to correct many-to-1
> join problem when patient has multiple surgeries in the dataset

R makes the same mistake loud:

```{r}
#| warning: true
#| eval: !expr requireNamespace("dplyr", quietly = TRUE)
a <- data.frame(id = c("A", "A"), x = 1:2)
b <- data.frame(id = c("A", "A"), y = 3:4)
nrow(dplyr::left_join(a, b, by = "id"))
```

Four rows from two — and `dplyr` says so, out loud, which SAS never did.

You can silence that warning with `relationship = "many-to-many"`, and there
are joins where the fan-out is genuinely intended. But reach for it only once
you have confirmed the duplication is what you want. Do not pass it reflexively
to quiet the output — that turns off the one check SAS never gave you.

### Dates use different origins

SAS counts days from 1960-01-01; R from 1970-01-01. The gap is 3,653 days.
`haven` converts correctly, so a value read through `snapshot_oracle()` is
already right. A hand-rolled conversion is where this goes wrong.

```{r}
as.Date(0, origin = "1970-01-01")   # R's zero
as.Date(0, origin = "1960-01-01")   # what a raw SAS number means
```

### Automatic variables do not exist in R

`_N_`, `_FREQ_`, and `_TYPE_` are created by SAS. In R you construct them, and
the name is not magic. Phase 0 of this migration found a defect of exactly this
class: `keep _freq_ tau` in one file versus `keep freq tau` in its sibling.
`_FREQ_` is the automatic variable; `freq` is an ordinary one that may not
exist. One of those programs produced wrong counts.

### `length 4` silently truncates

A SAS numeric written with `length 4` loses precision on disk. A value read
back from an old dataset may not equal a freshly computed R value, and **the
SAS side is the lossy one**. `tolerance` in `compare_built()` is what catches
this — but it is an **absolute** threshold, not a relative one. A single
absolute threshold cannot serve columns of mixed magnitude: a value large
enough to be at risk of `length 4` truncation (into the millions, for
identifiers or accumulated counts) can differ by more than a small absolute
`tolerance` while the difference is still storage precision, not a real
disagreement. `1234567` vs `1234567.06` reports `differs`, not
`within_tolerance`, with `max_abs_diff` of `0.06` — but `max_rel_diff` is
`4.9e-08`. For a large-magnitude variable, that tiny relative difference
alongside a `differs` verdict is the signal to suspect storage precision, not
the `within_tolerance` verdict itself. Read `max_rel_diff`, not the verdict
alone, when judging whether a difference is real.

## Verifying your own port

Build the variable in R, then compare against the oracle:

```{r}
oracle <- data.frame(ccfidu = c("A1", "A2", "A3"), age = c(65.5, 70.2, 58.0))
mine   <- data.frame(ccfidu = c("A1", "A2", "A3"), age = c(65.5, 70.2, 58.5))

compare_built(oracle, mine, id = "ccfidu")
```

There is no overall pass or fail, deliberately. Read the table, and record a
resolution for every variable that is not `identical`.
